---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
library(tidyverse)

activityData <- read.csv("activity.csv")
activityData$date <- as.Date(activityData$date, "%Y-%m-%d")
activityData <- tibble(activityData)
```

## What is mean total number of steps taken per day?

```{r}
stepSummary <- activityData %>% 
  group_by(date) %>% 
  summarise(totalSteps = sum(steps, na.rm = TRUE), .groups = "keep")
meanSteps <- mean(stepSummary$totalSteps)
medianSteps <- median(stepSummary$totalSteps)
hist(stepSummary$totalSteps, main = "Total steps per day", xlab = "Steps")
```

The mean number of steps taken per day was `r meanSteps`.

The median number of steps taken per day was `r medianSteps`.


## What is the average daily activity pattern?

```{r}
stepSummary <- activityData %>%
  group_by(interval) %>%
  summarise(meanSteps = mean(steps, na.rm = TRUE), .groups = "keep")
plot(stepSummary$interval, 
     stepSummary$meanSteps, 
     type = "l", 
     main = "Activity Pattern", 
     xlab = "Time Interval", 
     ylab = "Average Steps")
maxInterval <- stepSummary$interval[which.max(stepSummary$meanSteps)]
```
The 5 minute interval with the most steps was `r maxInterval`

## Imputing missing values

```{r}
missingRows <- sum(!complete.cases(activityData))
```
`r missingRows` entries are missing from the data.

To impute the missing values I will use the mean value for that interval as calculated in the last section.

```{r}
imputedActivityData <- activityData

for(i in 1:nrow(imputedActivityData)) {
  if(is.na(imputedActivityData[i, "steps"])) {
    missingIndex <-
      which(stepSummary$interval == imputedActivityData$interval[i], 
            arr.ind = TRUE)
    
    imputedActivityData[i, "steps"] <- 
      as.integer(stepSummary$meanSteps[missingIndex])
  }
}

stepSummary <- imputedActivityData %>% 
  group_by(date) %>%
  summarise(totalSteps = sum(steps, na.rm = TRUE), .groups = "keep")

meanSteps <- mean(stepSummary$totalSteps)
medianSteps <- median(stepSummary$totalSteps)
```

The mean number of steps of the imputed data set is `r format(meanSteps)`. Because we imputed the mean based on the interval's average, not the day's average, this value is expected to be different than the previous mean. 

The median of the imputed data set is `r format(medianSteps)`. By imputing using the interval mean we introduced the same value into the data set multiple times. This resulted in a change to the median.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
activityData <- activityData %>% 
  add_column(weekday = rep("Null", times = nrow(activityData)))
isWeekday = as.numeric(format.Date(activityData$date, "%u")) < 6
activityData$weekday[isWeekday] <- "weekday"
activityData$weekday[!isWeekday] <- "weekend"
activityData$weekday <- as.factor(activityData$weekday)

activityData <- activityData %>% 
  group_by(interval, weekday) %>%
  summarise(meanSteps = mean(steps, na.rm = TRUE), .groups = "keep")

ggplot(data = activityData, aes(interval, meanSteps)) + 
  geom_line() + 
  facet_grid(weekday ~.) +
  labs(y = "Mean Steps")
```


